---
title: "One-Way ANOVA"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    theme: default
    css: mlp2.css
    self-contained: true
    slide-number: true
    footer: "STA4173 - Biostatistics - Fall 2025"
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Introduction: Topics

- Previous: two groups, continuous outcome

- Now: more than two groups, continuous outcome

- One-way ANOVA
    - posthoc testing
    - assumptions

- Kruskal-Wallis
    - posthoc testing

## Introduction: Analysis of Variance

- We have previously discussed testing the difference between two groups.

    - What about when there are three or more groups? 

- We will use a method called **an**alysis **o**f **va**riance (ANOVA).

    - This method partitions the variance of the outcome into variance due to the groups and variance due to "other" factors.

- Fun fact: the two-sample *t*-test is a special case of ANOVA.

    - If you perform ANOVA when comparing two means, you will obtain the same results as the two-sample *t*-test.
    
## Introduction: Analysis of Variance {.smaller}

```{r}
#| echo: false
library(ssstats)
library(tidyverse)
library(ggpubr)
set.seed(187782)
set.seed(8651)
data <- tibble(group = c(rep("A", 50), c(rep("B", 50))),
               value = if_else(group == "A", rnorm(100, mean = 100, sd = 15), rnorm(100, mean = 100, sd = 15)))
```

- Fun fact: the two-sample *t*-test is a special case of ANOVA.

- Two-sample *t*-test:

```{r}
#| echo: false
data %>% independent_mean_HT(continuous = value,
                             grouping = group,
                             alternative = "two",
                             alpha = 0.05)
```

- In ANOVA,

```{r}
#| echo: false
data %>% one_way_ANOVA(continuous = value, grouping = group)
```

## One-Way ANOVA

- In one-way ANOVA, we are partitioning the variability in our outcome (SS<sub>Total</sub>) into two pieces:
    - Variability due to the group (SS<sub>Treatment</sub>),
    - Variablity due to "other factors" (SS<sub>Error</sub>).
        - Think of this like a "catch all" for other sources of error -- things we did not adjust for in our model.

## One-Way ANOVA: ANOVA Table

- The computations for ANOVA are more involved than what we've seen before.  

- An ANOVA table will be constructed in order to perform the hypothesis test.

| **Source** | **Sum of Squares** | **df** | **Mean Squares** | ***F*** |
|-|-|-|-|-|
Treatment | SS<sub>Trt</sub> | df<sub>Trt</sub> | MS<sub>Trt</sub> | *F*<sub>0</sub>
Error | SS<sub>E</sub> | df<sub>E</sub> | MS<sub>E</sub> |
Total | SS<sub>Tot</sub> | df<sub>Tot</sub> | | |

- Once this is put together, we can perform the hypothesis test.

    - Our test statistic is the $F_0$.

## One-Way ANOVA: (Hand) Computations

- Before we begin our computations, it would be helpful if we know

$$\bar{x}, \ \ n_i, \ \ \bar{x}_i, \ \ s_i^2$$ 

- where,
    - $\bar{x}$ is the overall mean,
    - $n_i$ is the sample size for group $i$,
    - $\bar{x}_i$ is the mean for group $i$, and
    - $s_i^2$ is the variance for group $i$

## One-Way ANOVA: (Hand) Computations

- We begin our computations with the sums of squares: 

$$
\begin{align*}
    \text{SS}_{\text{Trt}} &= \sum_{i=1}^k n_i(\bar{x}_i-\bar{x})^2 \\
    \text{SS}_{\text{E}} &= \sum_{i=1}^k (n_i-1)s_i^2 \\
    \text{SS}_{\text{Tot}} &= \text{SS}_{\text{Trt}} + \text{SS}_{\text{E}}
\end{align*} 
$$ 
    
## One-Way ANOVA: (Hand) Computations

- Each sum of squares has degrees of freedom: 

$$
\begin{align*}
    \text{df}_{\text{Trt}} &= k-1\\
    \text{df}_{\text{E}} &= n-k\\
    \text{df}_{\text{Tot}} &= n-1
\end{align*} 
$$ 
    
## One-Way ANOVA: (Hand) Computations

- Once we have the sum of squares and corresponding degrees of freedom, we can compute the mean squares.

- In the case of one-way ANOVA,
$$
\begin{align*}
    \text{MS}_{\text{Trt}} &= \frac{\text{SS}_{\text{Trt}}}{\text{df}_{\text{Trt}}} \\
    \text{MS}_{\text{E}} &= \frac{\text{SS}_{\text{E}}}{\text{df}_{\text{E}}}
\end{align*}
$$
    - Note that there is **no** $\text{MS}_{\text{Tot}}$! 
    
## One-Way ANOVA: (Hand) Computations

- A note about mean squares: they are almost always constructed the same way.

$$\text{MS}_X = \frac{\text{SS}_X}{\text{df}_E}$$

- This is important to know for future statistics courses that may have you calculating things by hand. 

## One-Way ANOVA: (Hand) Computations

- Finally, we have the test statistic. 

- Generally, we construct an $F$ for ANOVA by dividing the MS of interest by MS_E_,

$$F_X = \frac{\text{MS}_X}{\text{MS}_{\text{E}}}$$

- In one-way ANOVA, we are only constructing the $F$ for treatment,

$$F_0 = \frac{\text{MS}_{\text{Trt}}}{\text{MS}_{\text{E}}} $$

## One-Way ANOVA: (Hand) Computations

- We are finally done constructing our ANOVA table! As a reminder,

| **Source** | **Sum of Squares** | **df** | **Mean Squares** | ***F*** |
|-|-|-|-|-|
Treatment | SS<sub>Trt</sub> | df<sub>Trt</sub> | MS<sub>Trt</sub> | *F*<sub>0</sub>
Error | SS<sub>E</sub> | df<sub>E</sub> | MS<sub>E</sub> |
Total | SS<sub>Tot</sub> | df<sub>Tot</sub> | | |    

## One-Way ANOVA: ANOVA Table (R)

- We will use the `one_way_ANOVA_table()` function from `library(ssstats)` to construct the ANOVA table.

```{r}
#| eval: false
dataset_name %>% one_way_ANOVA_table(continuous = continuous_variable,
                                     grouping = grouping_variable)
```

## One-Way ANOVA: ANOVA Table

- In the magical land of Equestria, ponies come in different types: Unicorns, Pegasi, Earth Ponies, and Alicorns. While each group has unique abilities, some researchers in Twilight Sparkle's lab are curious whether these pony types differ in their overall magic ability scores, a standardized measure that combines magical potential, control, and versatility.

- To investigate this, they collect data (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). The researchers want to know if there is difference in magic ability scores (*ability_score*) among the four pony types. We will test at the $\alpha=0.05$ level.

- Let's first create the ANOVA table. How should we update this code?

```{r}
#| eval: false
dataset_name %>% one_way_ANOVA_table(continuous = continuous_variable,
                                     grouping = grouping_variable)
```

```{r}
#| echo: false

library(tidyverse)

set.seed(673560)  # for reproducibility

# Simulate data for each pony type
magical_studies <- tibble(pony_type = rep(c("Unicorn", "Pegasus", "Earth", "Alicorn"), each = 35),
                          ability_score = c(rnorm(35, mean = 75, sd = 10),
                                            rnorm(35, mean = 74, sd = 10),
                                            rnorm(35, mean = 76, sd = 10),
                                            rnorm(35, mean = 75, sd = 10)),
                          coordination_score = c(rnorm(35, mean = 85, sd = 8),
                                                 rnorm(35, mean = 78, sd = 8), 
                                                 rnorm(35, mean = 73, sd = 8), 
                                                 rnorm(35, mean = 88, sd = 8)))
```

## One-Way ANOVA: ANOVA Table

- In the magical land of Equestria, ponies come in different types: Unicorns, Pegasi, Earth Ponies, and Alicorns. While each group has unique abilities, some researchers in Twilight Sparkle's lab are curious whether these pony types differ in their overall magic ability scores, a standardized measure that combines magical potential, control, and versatility.

- To investigate this, they collect data (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). The researchers want to know if there is difference in magic ability scores (*ability_score*) among the four pony types.

- Let's first create the ANOVA table. Our updated code:

```{r}
#| eval: false
magical_studies %>% one_way_ANOVA_table(continuous = ability_score,
                                        grouping = pony_type)
```

## One-Way ANOVA: ANOVA Table

- Running the code,

```{r}
magical_studies %>% one_way_ANOVA_table(continuous = ability_score,
                                        grouping = pony_type)
```


## Hypothesis Testing: One-Way-ANOVA

- In one-way ANOVA, hypotheses always take the same form:

    - $H_0: \ \mu_1 = \mu_2 = ... = \mu_k$
    - $H_1:$ at least one is different

- Note: you **must** fill in the "k" when writing hypotheses!
    - e.g., if there are four means, your hypotheses are
        - $H_0: \ \mu_1 = \mu_2 = \mu_3 = \mu_4$
        - $H_1:$ at least one is different
    - e.g., in our MLP example,
        - $H_0: \ \mu_{\text{unicorn}} = \mu_{\text{earth}} = \mu_{\text{alicorn}} = \mu_{\text{pegasus}}$
        - $H_1:$ at least one is different

## Hypothesis Testing: One-Way ANOVA

**Test statistic:**

$$
F_0 = \frac{\text{MS}_{\text{Trt}}}{\text{MS}_{\text{E}}}
$$


***p*-Value:**

$$
p = P[F_{k-1,n-k} \ge F_0]
$$


## Hypothesis Testing: One-Way ANOVA (R)

- We will use the `one_way_ANOVA()` function from `library(ssstats)` to construct the corresponding hypothesis test.

```{r}
#| eval: false
dataset_name %>% one_way_ANOVA(continuous = continuous_variable,
                               grouping = grouping_variable,
                               alpha = specified_alpha)
```

## Hypothesis Testing: One-Way ANOVA

- In the magical land of Equestria, ponies come in different types: Unicorns, Pegasi, Earth Ponies, and Alicorns. While each group has unique abilities, some researchers in Twilight Sparkle's lab are curious whether these pony types differ in their overall magic ability scores, a standardized measure that combines magical potential, control, and versatility.

- To investigate this, they collect data (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). The researchers want to know if there is difference in magic ability scores (*ability_score*) among the four pony types. We will test at the $\alpha=0.05$ level.

- Let's now formulate the hypothesis test. How should we update this code?

```{r}
#| eval: false
dataset_name %>% one_way_ANOVA(continuous = continuous_variable,
                               grouping = grouping_variable,
                               alpha = specified_alpha)
```

## Hypothesis Testing: One-Way ANOVA

- In the magical land of Equestria, ponies come in different types: Unicorns, Pegasi, Earth Ponies, and Alicorns. While each group has unique abilities, some researchers in Twilight Sparkle's lab are curious whether these pony types differ in their overall magic ability scores, a standardized measure that combines magical potential, control, and versatility.

- To investigate this, they collect data (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). The researchers want to know if there is difference in magic ability scores (*ability_score*) among the four pony types. We will test at the $\alpha=0.05$ level.

- Let's now formulate the hypothesis test. Our updated code,

```{r}
#| eval: false
magical_studies %>% one_way_ANOVA(continuous = ability_score,
                                  grouping = pony_type, 
                                  alpha = 0.05)
```

## Hypothesis Testing: One-Way ANOVA

- Running the code,

```{r}
magical_studies %>% one_way_ANOVA(continuous = ability_score,
                                  grouping = pony_type, 
                                  alpha = 0.05)
```

## Hypothesis Testing: One-Way ANOVA

- **Hypotheses**
    - $H_0: \ \mu_{\text{alicorn}} = \mu_{\text{unicorn}} = \mu_{\text{earth}} = \mu_{\text{pegasus}}$
    - $H_1:$ at least one $\mu_i$ is different
  
- **Test Statistic and *p*-Value**
    - $F_0 = 1.526$ 
    - $p = 0.210$

- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha=0.05$.
  
- **Conclusion/Interpretation**
    - Fail to reject $H_0$ ($p \text{ vs } \alpha \to p = 0.210 > 0.05$). There is not sufficient evidence to suggest that there is a difference in magical ability between the types of ponies.
    
## Hypothesis Testing: One-Way ANOVA

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know if there is difference in magical coordination scores (*coordination_score*) among the four pony types. We will again test at the $\alpha=0.05$ level.

- Let's first create the ANOVA table. How should we update this code? 

```{r}
#| eval: false
dataset_name %>% one_way_ANOVA_table(continuous = continuous_variable,
                                     grouping = grouping_variable)
```

## Hypothesis Testing: One-Way ANOVA

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know if there is difference in magical coordination scores (*coordination_score*) among the four pony types. We will again test at the $\alpha=0.05$ level.

- Let's first create the ANOVA table. Our updated code:

```{r}
#| eval: false
magical_studies %>% one_way_ANOVA_table(continuous = coordination_score,
                                        grouping = pony_type)
```

## Hypothesis Testing: One-Way ANOVA

- Running the code:

```{r}
magical_studies %>% one_way_ANOVA_table(continuous = coordination_score,
                                        grouping = pony_type)
```

## Hypothesis Testing: One-Way ANOVA

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know if there is difference in magical coordination scores (*coordination_score*) among the four pony types. We will again test at the $\alpha=0.05$ level.

- Let's now formulate the hypothesis test. How should we update this code?

```{r}
#| eval: false
dataset_name %>% one_way_ANOVA(continuous = continuous_variable,
                               grouping = grouping_variable,
                               alpha = specified_alpha)
```

## Hypothesis Testing: One-Way ANOVA

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know if there is difference in magical coordination scores (*coordination_score*) among the four pony types. We will again test at the $\alpha=0.05$ level.

- Let's now formulate the hypothesis test. Our updated code:

```{r}
#| eval: false
magical_studies %>% one_way_ANOVA(continuous = coordination_score,
                                  grouping = pony_type,
                                  alpha = 0.05)
```

## Hypothesis Testing: One-Way ANOVA

- Running the code:

```{r}
magical_studies %>% one_way_ANOVA(continuous = coordination_score,
                                  grouping = pony_type,
                                  alpha = 0.05)
```

## Hypothesis Testing: One-Way ANOVA

- **Hypotheses**
    - $H_0: \ \mu_{\text{alicorn}} = \mu_{\text{unicorn}} = \mu_{\text{earth}} = \mu_{\text{pegasus}}$
    - $H_1:$ at least one $\mu_i$ is different
  
- **Test Statistic and *p*-Value**
    - $F_0 = 18.396$ 
    - $p < 0.001$

- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha=0.05$.
  
- **Conclusion/Interpretation**
    - Reject $H_0$ ($p \text{ vs } \alpha \to p < 0.001 < 0.05$). There is sufficient evidence to suggest that there is a difference in magical coordination between the types of ponies.

## Introduction: Posthoc Testing

- There's a difference, but what is the difference? 

- Let's look at summary statistics:

```{r}
magical_studies %>% 
  group_by(pony_type) %>%
  mean_median(coordination_score)
```

- Can we definitively say which groups are different...?

## Introduction: Posthoc Testing

- Recall our hypotheses in one-way ANOVA,

    - $H_0: \mu_1 = \mu_2 = ... = \mu_k$
    - $H_1:$ at least one $\mu_i$ is different
    
- The $F$ test does not tell us *which* mean is different... only that a difference exists.

- In theory, we could perform repeated $t$ tests to determine pairwise differences.
    - Recall that ANOVA is an extension of the $t$ test... or that the $t$ test is a special case of ANOVA.
    - However, this will increase the Type I error rate ($\alpha$).

## Introduction: Posthoc Testing

- Recall that the Type I error rate, $\alpha$, is the probability of *incorrectly* rejecting $H_0$.

    - i.e., we are saying there is a difference between the means when there is actually *not* a difference.
    
- Suppose we are comparing 5 groups.

    - This is 10 pairwise comparisons!! 
        
        - 1-2, 1-3, 1-4, 1-5, 2-3, 2-4, 2-5, 3-4, 3-5, 4-5
        
    - If we perform repeated $t$ tests under $\alpha=0.05$, we are inflating the Type I error to 0.40! 😵

## Introduction: Posthoc Testing

- When performing posthoc comparisons, we can choose one of two paths:

    - Control the Type I (familywise) error rate.
    - Do not control the Type I error rate.
    
- Note that controlling the Type I error rate is more conservative than when we do not control it.

    - "Conservative" = more difficult to reject.
    
- Generally, statisticians:

    - *do not* control the Type I error rate if examining the results of pilot/preliminary studies that are exploring for general relationships.
    
    - *do* control the Type I error rate if examining the results of confirmatory studies and are attempting to confirm relationships observed in pilot/preliminary studies.
    
## Introduction: Posthoc Testing

- The posthoc tests we will learn:

    - Tukey's test
    
        - Performs all pairwise tests and controls the Type I error rate
        
    - Fisher's least significant difference 
    
        - Performs all pairwise tests but does not control the Type I error rate
        
- **Caution**: we should *only* perform posthoc tests if we have determined that a general difference exists!

    - i.e., when we reject when looking at the $F$ test in ANOVA
    
## Posthoc Testing: Tukey's Test

- Tukey's test allows us to do all pairwise comparisons while controlling $\alpha$.

- **Hypotheses**
    - $H_0: \ \mu_i = \mu_j$ 
    - $H_1: \ \mu_i \ne \mu_j$
    
- **Test Statistic**

$$
Q = \frac{|\bar{y}_i - \bar{y}_j|}{\sqrt{ \frac{\text{MS}_{\text{E}}}{2} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$

## Posthoc Testing: Tukey's Test (R)

- We will use the `posthoc_tukey()` function from `library(ssstats)` to perform Tukey's posthoc test (resulting $p$ are adjusted for multiple comparisons).

```{r}
#| eval: false
dataset_name %>% posthoc_tukey(continuous = continuous_variable,
                               grouping = grouping_variable)
```

## Posthoc Testing: Tukey's Test

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know what differences exist in their magical coordination scores (*coordination_score*). We will again test at the $\alpha=0.05$ level.

- Let's now formulate Tukey's posthoc test. How should we update this code?

```{r}
#| eval: false
dataset_name %>% posthoc_tukey(continuous = continuous_variable,
                               grouping = grouping_variable)
```

## Posthoc Testing: Tukey's Test

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know what differences exist in their magical coordination scores (*coordination_score*). We will again test at the $\alpha=0.05$ level.

- Let's now formulate Tukey's posthoc test. Our updated code:

```{r}
#| eval: false
magical_studies %>% posthoc_tukey(continuous = coordination_score,
                                  grouping = pony_type)
```

## Posthoc Testing: Tukey's Test

- Running the code,

```{r}
magical_studies %>% posthoc_tukey(continuous = coordination_score,
                                  grouping = pony_type)
```

## Posthoc Testing: Tukey's Test

- Restating results,
    - $\mu_{\text{alicorn}} \ne \mu_{\text{earth}}$ ($\bar{x}_d = 12.63$, $p < 0.001$)
    - $\mu_{\text{alicorn}} \ne \mu_{\text{pegasus}}$ ($\bar{x}_d = 10.18$, $p < 0.001$)
    - $\mu_{\text{unicorn}} \ne \mu_{\text{earth}}$ ($\bar{x}_d = 9.83$, $p < 0.001$)
    - $\mu_{\text{unicorn}} \ne \mu_{\text{pegasus}}$ ($\bar{x}_d = 7.38$, $p = 0.001$)
    - $\mu_{\text{alicorn}} = \mu_{\text{unicorn}}$ ($\bar{x}_d = 2.80$, $p = 0.487$)
    - $\mu_{\text{pegasus}} = \mu_{\text{earth}}$ ($\bar{x}_d = 2.44$, $p = 0.602$)

## Posthoc Testing: Fisher's Test

- Fisher's allows us to test all pairwise comparisons but *does not* control the $\alpha$.

- **Hypotheses:**
    - $H_0: \ \mu_i = \mu_j$
    - $H_1: \ \mu_i \ne \mu_j$

- **Test Statistic:**

$$
t = \frac{|\bar{y}_i - \bar{y}_j|}{\sqrt{ \text{MS}_{\text{E}} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$

## Posthoc Testing: Fisher's Test (R)

- We will use the `posthoc_fisher()` function from `library(ssstats)` to perform Tukey's posthoc test (resulting $p$ are *not* adjusted for multiple comparisons).

```{r}
#| eval: false
dataset_name %>% posthoc_fisher(continuous = continuous_variable,
                                grouping = grouping_variable)
```

## Posthoc Testing: Fisher's Test 

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know what differences exist in their magical coordination scores (*coordination_score*). We will again test at the $\alpha=0.05$ level.

- Let's now formulate Fisher's posthoc test. How should we update this code?

```{r}
#| eval: false
dataset_name %>% posthoc_tukey(continuous = continuous_variable,
                               grouping = grouping_variable)
```

## Posthoc Testing: Fisher's Test

- Twilight Sparkle is now leading a study to understand whether ponies from different types specialize in different areas of magical strength. She develops a "magical coordination" score, which reflects how well a pony can use magic to interact with objects in motion (e.g., catching falling books).

- To investigate this, data is collected (***magical_studies***) on a random sample of ponies from each pony type (*pony_type*). Twilight Sparkle wants to know what differences exist in their magical coordination scores (*coordination_score*). We will again test at the $\alpha=0.05$ level.

- Let's now formulate Fisher's posthoc test. Our updated code:

```{r}
#| eval: false
magical_studies %>% posthoc_fisher(continuous = coordination_score,
                                   grouping = pony_type)
```

## Posthoc Testing: Fisher's Test 

- Running the code,

```{r}
magical_studies %>% posthoc_fisher(continuous = coordination_score,
                                   grouping = pony_type)
```

## Posthoc Testing: Fisher's Test

- Restating results,
    - $\mu_{\text{alicorn}} \ne \mu_{\text{earth}}$ ($\bar{x}_d = 12.63$, $p < 0.001$)
    - $\mu_{\text{alicorn}} \ne \mu_{\text{pegasus}}$ ($\bar{x}_d = 10.18$, $p < 0.001$)
    - $\mu_{\text{unicorn}} \ne \mu_{\text{earth}}$ ($\bar{x}_d = 9.83$, $p < 0.001$)
    - $\mu_{\text{unicorn}} \ne \mu_{\text{pegasus}}$ ($\bar{x}_d = 7.38$, $p < 0.001$)
    - $\mu_{\text{alicorn}} = \mu_{\text{unicorn}}$ ($\bar{x}_d = 2.80$, $p = 0.151$)
    - $\mu_{\text{pegasus}} = \mu_{\text{earth}}$ ($\bar{x}_d = 2.44$, $p = 0.227$)

## Posthoc Testing: Tukey's vs. Fisher's

- We have now learned:
    - Tukey's test: Performs all pairwise tests and controls the Type I error rate
    - Fisher's test: Performs all pairwise tests but does not control the Type I error rate

- Sometimes Tukey's and Fisher's will agree with each other.
    - This is the case in our example

- Other times they do not agree -- by design, Tukey's makes it *harder* to reject.
    - We may see more rejections when using Fisher's

## Posthoc Testing: Tukey's vs. Fisher's

- Comparing the two side-by-side,

| Pairwise Comparison    | $\bar{x}_d$ | Unadjusted $p$ | Adjusted $p$ |
|------------------------|-------------:|----------------:|--------------:|
| Alicorn vs. Earth      | 12.63       | < 0.001       | < 0.001     |
| Alicorn vs. Pegasus    | 10.18       | < 0.001       | < 0.001     |
| Unicorn vs. Earth      | 9.83        | < 0.001       | < 0.001     |
| Unicorn vs. Pegasus    | 7.38        | < 0.001       | 0.001       |
| Alicorn vs. Unicorn    | 2.80        | 0.151         | 0.487       |
| Pegasus vs. Earth      | 2.44        | 0.227         | 0.602       |



## Wrap Up

- Today we have covered one-way ANOVA.
    - One-way ANOVA table
    - Test for equality among $k$ means
    - Posthoc testing (pairwise comparisons)
        - Tukey's (ANOVA, ajdusted)
        - Fisher's (ANOVA, unadjusted)

## Wrap Up

- Upcoming:
    - M/T:
        - ANOVA assumptions
        - Nonparametric alternative to one-way ANOVA (Kruskal-Wallis)
        - Nonparametric alternative to posthoc testing (Dunn's test)
    - W/R:
        - Lab: one-way ANOVA and Kruskal-Wallis
        - Quiz: one-way ANOVA and Kruskal-Wallis
    
    


