---
title: "Module 2 Review"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    theme: default
    css: mlp2.css
    self-contained: true
    slide-number: true
    footer: "STA4173 - Biostatistics - Fall 2025"
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Putting It All Together

```{r}
#| echo: false
library(tidyverse)
library(ssstats)
```

- One-way ANOVA / Kruskal-Wallis:
    - Continuous outcome.
    - Only one factor (grouping variable).
        - More than two groups.
    
- Two-way ANOVA:
    - Continuous outcome.
    - Two factors (grouping variables), A & B.
        - More than two groups in each factor.
    - Interaction between factors A & B.

- ANOVA assumptions:
    - Normality of residuals.
    - Equal variances.
    
## Example 1

```{r}
#| echo: false
set.seed(42561) 
library(tidyverse)
means <- tibble(
  fertilizer = c("Nook's GreenGro", "Leif's Compost Tea", "Brewster's Grounds", "No Fertilizer"),
  mu         = c(1.8,                2.2,                 2.4,                  1.6),
  sigma      = 0.30)

n_per_group <- 20

veg <- means %>%
  group_by(fertilizer) %>%
  summarize(
    yield_kg = rnorm(n_per_group, mean = mu, sd = sigma),
    .groups = "drop_last"
  ) %>%
  ungroup() %>%
  mutate(fertilizer = factor(fertilizer,
                             levels = c("No Fertilizer","Nook's GreenGro","Leif's Compost Tea","Brewster's Grounds")))
```

- Leif has started a veggie-gardening club in Animal Crossing. Four popular fertilizer options are being tried on carrot plots:
    1. Nook's GreenGro
    2. Leif's Compost Tea
    3. Brewster's Grounds
    4. No fertilizer (control)

- Each option is applied to several 1 m^2^ carrot plots (one fertilizer per plot). At harvest, villagers record the carrot yield (kg) from each plot. 

- The research question is, *do mean carrot yields differ among the four fertilizer options*?

## Example 1

- Looking at summary statistics,

```{r}
#| echo: false
veg %>% group_by(fertilizer) %>%
  mean_median(yield_kg)
```

## Example 1

- Checking the ANOVA assumptions,

```{r}
#| fig-align: 'center'
#| echo: false
veg %>% ANOVA_assumptions(continuous = yield_kg,
                          group = fertilizer)
```

- Normality (qq plots) looks okay.

- Variance looks okay. 

## Example 1

- Checking the ANOVA assumptions,

```{r}
#| fig-align: 'center'
#| echo: false
veg %>% variances_HT(continuous = yield_kg,
                          group = fertilizer)
```

- Is the variance assumption broken?

## Example 1

- Constructing the one-way ANOVA test,

```{r}
#| echo: false
veg %>% one_way_ANOVA(continuous = yield_kg,
                      group = fertilizer)
```

- Is there a significant difference between the fertilizers?

## Example 1

- Suppose we want to look at posthoc tests to see which fertilizers differ.

- If we want to adjust, we will use Tukey's,

```{r}
#| echo: false
veg %>% posthoc_tukey(continuous = yield_kg,
                        group = fertilizer)
```

- Which fertilizers differ significantly?

## Example 1

- Suppose we want to look at posthoc tests to see which fertilizers differ.

- If we want to adjust, we will use Fisher's,

```{r}
#| echo: false
veg %>% posthoc_fisher(continuous = yield_kg,
                        group = fertilizer)
```

- Which fertilizers differ significantly?

## Example 2

- Leif notices some strange patterns this season. We now are testing our four fertilizer options  on 1 m^2^ tomato plots:
    1. Nook's GreenGro
    2. Leif's Compost Tea
    3. Brewster's Grounds
    4. No fertilizer (control)

- Each option is applied to several 1 m^2^ tomato plots (one fertilizer per plot). At harvest, villagers record the tomato yield (kg) from each plot. 

- The research question is, *do mean tomato yields differ among the four fertilizer options*?

## Example 2

- Looking at summary statistics, 

```{r}
#| echo: false

set.seed(777)
n <- 18

# helper to set lognormal so mean ≈ target_mu
r_lognorm_mean <- function(n, target_mu, sdlog){
  meanlog <- log(target_mu) - 0.5*sdlog^2
  rlnorm(n, meanlog = meanlog, sdlog = sdlog)
}

veg2 <- bind_rows(
  tibble(fertilizer = "No Fertilizer",
         yield_kg   = rnorm(n, mean = 1.6, sd = 0.25)),
  tibble(fertilizer = "Nook's GreenGro",
         yield_kg   = rnorm(n, mean = 1.9, sd = 0.30)),
  tibble(fertilizer = "Leif's Compost Tea",
         # skew: log-normal with same rough mean as 2.2
         yield_kg   = r_lognorm_mean(n, target_mu = 2.2, sdlog = 0.35)),
  tibble(fertilizer = "Brewster's Grounds",
         # heteroskedastic: much larger variance
         yield_kg   = rnorm(n, mean = 2.4, sd = 0.60))
) %>%
  mutate(fertilizer = factor(fertilizer,
    levels = c("No Fertilizer","Nook's GreenGro",
               "Leif's Compost Tea","Brewster's Grounds")))


veg2 %>% group_by(fertilizer) %>%
  mean_median(yield_kg)
```

## Example 2

- Checking the ANOVA assumptions,

```{r}
#| fig-align: 'center'
#| echo: false
veg2 %>% ANOVA_assumptions(continuous = yield_kg,
                          group = fertilizer)
```

- Normality (qq plots) looks okay...?

- Variance is questionable...

## Example 2

- Checking the ANOVA assumptions,

```{r}
#| fig-align: 'center'
#| echo: false
veg2 %>% variances_HT(continuous = yield_kg,
                          group = fertilizer)
```

- Is the variance assumption broken?

## Example 2

- Constructing the Kruskal-Wallis test,

```{r}
#| echo: false
veg2 %>% kruskal_HT(continuous = yield_kg,
                      group = fertilizer)
```

- Is there a significant difference between the fertilizers?

## Example 2

- Suppose we want to look at posthoc tests to see which fertilizers differ.

- If we want to adjust, we will use the adjusted version of Dunn's test,

```{r}
#| echo: false
veg2 %>% posthoc_dunn(continuous = yield_kg,
                        group = fertilizer, 
                      adjust = TRUE)
```

- Which fertilizers differ significantly?

## Example 2

- Suppose we want to look at posthoc tests to see which fertilizers differ.

- If we do not want to adjust, we will use the unadjusted version of Dunn's test,

```{r}
#| echo: false
veg2 %>% posthoc_dunn(continuous = yield_kg,
                        group = fertilizer, 
                      adjust = FALSE)
```

- Which fertilizers differ significantly?


## Example 3

- Leif is now testing the four fertilizer options on both carrot and tomato plots:
    1. Nook's GreenGro
    2. Leif's Compost Tea
    3. Brewster's Grounds
    4. No fertilizer (control)
	  
- The research questions are:
    1. *Is there an interaction between fertilizer type and crop type on yield?*
    2. *If no interaction, do mean crop yields differ among the four fertilizer options?*
    3. *If no interaction, do mean crop yields differ between carrot and tomato plots?*
    
## Example 3    
	  
- Looking at summary statistics,	  

```{r}
#| echo: false

set.seed(4242)
library(tidyverse)

# design parameters
n <- 15
fertilizers <- c("No Fertilizer", "Nook's GreenGro",
                 "Leif's Compost Tea", "Brewster's Grounds")
crops <- c("Carrot", "Tomato")

# define mean structure (interaction allowed)
# overall pattern: fertilizers improve yield, slightly more so for tomatoes
mean_matrix <- tribble(
  ~fertilizer,          ~Carrot, ~Tomato,
  "No Fertilizer",       1.6,     1.7,
  "Nook's GreenGro",     1.8,     1.9,
  "Leif's Compost Tea",  2.2,     2.3,
  "Brewster's Grounds",  2.4,     2.5
)

sigma <- 0.30  # equal SD across all cells

# generate tidy dataset
veg_norm <- mean_matrix %>%
  pivot_longer(-fertilizer, names_to="crop", values_to="mu") %>%
  group_by(fertilizer, crop) %>%
  summarise(
    yield_kg = rnorm(n, mean = mu, sd = sigma),
    .groups = "drop"
  ) %>%
  mutate(
    fertilizer = factor(fertilizer, levels = fertilizers),
    crop = factor(crop, levels = crops)
  )

veg_norm %>% group_by(crop, fertilizer) %>%
  mean_median(yield_kg)
```

## Example 3

- Checking the two-way ANOVA assumptions,

```{r}
#| echo: false
#| fig-align: 'center'
veg_norm %>% ANOVA2_assumptions(continuous = yield_kg,
                           A = fertilizer,
                           B = crop)
```

## Example 3

- Examining the two-way ANOVA (remember to always check for the interaction first!),

```{r}
#| echo: false
veg_norm %>% two_way_ANOVA_table(continuous = yield_kg,
                           A = fertilizer,
                           B = crop)
```

- Is the interaction term significant? 

- Should we keep or remove the interaction term?

## Example 3

- Removing the interaction term,

```{r}
#| echo: false
veg_norm %>% two_way_ANOVA_table(continuous = yield_kg,
                           A = fertilizer,
                           B = crop,
                           interaction = FALSE)
```

- Are there signfiicant main effects?

## Example 3

- Looking at the tests for main effects,

```{r}
#| echo: false
veg_norm %>% two_way_ANOVA(continuous = yield_kg,
                           A = fertilizer,
                           B = crop,
                           interaction = FALSE)
```

- Is fertilizer a significant factor for crop yield?

- Is crop type a significant factor for crop yield?

## Example 3

- Looking at the profile plot,

```{r}
#| echo: false
#| fig-align: 'center'
veg_norm %>% profile_plot(continuous = yield_kg,
                           xaxis = fertilizer,
                           lines = crop) + 
  labs(color = "Crop Type",
       x = "Fertilizer Type",
       y = "Average Yield (kg)")
```

- What statements can we make about this analysis?

## Example 4

- Leif is now testing the four fertilizer options on both pumpkin, wheat, and potato plots:
    1. Nook's GreenGro
    2. Leif's Compost Tea
    3. Brewster's Grounds
    4. No fertilizer (control)

- The research questions are:
    1. *Is there an interaction between fertilizer type and crop type on yield?*
    2. *If no interaction, do mean crop yields differ among the four fertilizer options?*
    3. *If no interaction, do mean crop yields differ among the three crop types?*

## Example 4

- Looking at summary statistics,	  

```{r}
#| echo: false

set.seed(810)

n <- 12
fert_levels <- c("No Fertilizer","Nook's GreenGro","Leif's Compost Tea","Brewster's Grounds")
crop_levels <- c("Pumpkin","Wheat","Potato")
sigma <- 0.30

mean_table <- tribble(
  ~crop,     ~fertilizer,           ~mu,
  "Pumpkin", "No Fertilizer",        2.0,
  "Pumpkin", "Nook's GreenGro",      2.2,
  "Pumpkin", "Leif's Compost Tea",   2.5,
  "Pumpkin", "Brewster's Grounds",   2.8,
  "Wheat",   "No Fertilizer",        1.6,
  "Wheat",   "Nook's GreenGro",      2.0,
  "Wheat",   "Leif's Compost Tea",   2.4,
  "Wheat",   "Brewster's Grounds",   2.3,
  "Potato",  "No Fertilizer",        2.1,
  "Potato",  "Nook's GreenGro",      2.3,
  "Potato",  "Leif's Compost Tea",   2.4,
  "Potato",  "Brewster's Grounds",   2.4
)

veg_int <- mean_table %>%
  group_by(crop, fertilizer) %>%
  summarise(yield_kg = rnorm(n, mean = mu, sd = sigma), .groups="drop") %>%
  mutate(
    fertilizer = factor(fertilizer, levels = fert_levels),
    crop = factor(crop, levels = crop_levels)
  )

veg_int %>% group_by(crop, fertilizer) %>%
  mean_median(yield_kg)
```

## Example 4

- Checking the two-way ANOVA assumptions,

```{r}
#| echo: false
#| fig-align: 'center'
veg_int %>% ANOVA2_assumptions(continuous = yield_kg,
                           A = fertilizer,
                           B = crop)
```

## Example 4

- Examining the two-way ANOVA (remember to always check for the interaction first!),

```{r}
#| echo: false
veg_int %>% two_way_ANOVA_table(continuous = yield_kg,
                           A = fertilizer,
                           B = crop)
```

- Is the interaction term significant? 

- Should we keep or remove the interaction term?

## Example 4

- Looking at the test for the interaction,

```{r}
#| echo: false
veg_int %>% two_way_ANOVA(continuous = yield_kg,
                           A = fertilizer,
                           B = crop,
                           interaction = TRUE)
```

- Is there a significant interaction?

## Example 4

- Looking at the profile plot,

```{r}
#| echo: false
#| fig-align: 'center'
veg_int %>% profile_plot(continuous = yield_kg,
                           xaxis = fertilizer,
                           lines = crop) + 
  labs(color = "Crop Type",
       x = "Fertilizer Type",
       y = "Average Yield (kg)")
```

- What statements can we make about this analysis?


## Wrap Up        

- Today's review:
    - One-way ANOVA
    - Kruskal-Wallis
    - Posthoc testing
    - Two-way ANOVA
    - Interaction terms
    - Profile plots
    - ANOVA Assumptions
    
- Next class:   
    - Regression!!!!! 💕
