---
title: "Module 3 Review"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    theme: default
    css: mlp2.css
    self-contained: true
    slide-number: true
    footer: "STA4173 - Biostatistics - Fall 2025"
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Putting It All Together

```{r}
#| echo: false
library(tidyverse)
library(ssstats)
```

- Relationships between continuous variables.

- Correlation: 
    - Unitless measure of linear (Pearson's) or monotonic (Spearman's) <u>pairwise</u> relationships.
        - Does not *adjust* for other variables.
    - Quantifies strength and direction of relationship.
        - Allows us to compare strength of relationships across different pairs of variables.
        
- Linear regression:
    - Models relationship between continuous outcome and one *or more* predictors.
        - Units are attached, so the relationship can be stated in context.
    - Provides estimates of association *adjusted* for other variables in the model.
    - Allows for prediction of outcome based on predictor values.
    
```{r}
#| echo: false
set.seed(98165)
library(tidyverse)
library(broom)
library(ssstats)
library(ggpubr)

puzzles <- tibble(
  # continuous predictors
  MindPractice    = pmax(0, rnorm(150, mean = 2.8, sd = 1.1)),   # hours
  SleepHours      = pmin(pmax(rnorm(150, mean = 7.2, sd = 1.2), 3), 10),  # hours, clipped to 3-10
  FocusMinutes    = pmax(0, rnorm(150, mean = 10, sd = 5)),      # minutes
  PuzzleComplexity= pmin(pmax(rnorm(150, mean = 6, sd = 2), 1), 10)       # 1-10 scale
) |>
  # linear data-generating process: helpful skills reduce time; complexity increases time
  mutate(
    SolveTime = 35
      - 4.0 * MindPractice
      - 1.3 * SleepHours
      - 0.05 * FocusMinutes
      + 2.8 * PuzzleComplexity
      + rnorm(150, 0, 5.5)
  ) |>
  mutate(SolveTime = pmax(SolveTime, 2))  # keep times sensible
```    
        
## Example 

- During a puzzle-heavy session in Ponyville, Twilight wants to understand what helps small teams solve logic puzzles faster. For each puzzle, she records:

    - *MindPractice*: hours of individual "mind" drills the day before.
    - *SleepHours*: hours slept the night before.
    - *FocusMinutes*: minutes of calm focus/box-breathing right before starting.
    - *PuzzleComplexity*: a continuous difficulty score Twilight assigns from 1–10.
    - *SolveTime* — minutes to solve the puzzle (lower is better).

- Our task is to determine how MindPractice, SleepHours, FocusMinutes, and PuzzleComplexity relate to SolveTime.

## Example 

- Let's first look at pairwise scatterplots.

```{r}
#| echo: false
#| fig-align: 'center'

p1 <- puzzles %>% ggplot(aes(y = SolveTime)) +
  geom_point(aes(x = MindPractice)) +
  theme_bw() +
  labs(x = "Mind Practice (hours)", 
       y = "Solve Time (minutes)")

p2 <- puzzles %>% ggplot(aes(y = SolveTime)) +
  geom_point(aes(x = SleepHours)) +
  theme_bw() +
  labs(x = "Amount of Sleep (hours)", 
       y = "Solve Time (minutes)")

p3 <- puzzles %>% ggplot(aes(y = SolveTime)) +
  geom_point(aes(x = FocusMinutes)) +
  theme_bw() +
  labs(x = "Focus Time (minutes)", 
       y = "Solve Time (minutes)")

p4 <- puzzles %>% ggplot(aes(y = SolveTime)) +
  geom_point(aes(x = PuzzleComplexity)) +
  theme_bw() +
  labs(x = "Puzzle Complexity (rating)", 
       y = "Solve Time (minutes)")

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2)

```

## Example 

- Next, let's determine what method of correlation we should use. Checking normality,

```{r}
#| echo: false
#| fig-align: 'center'
p1 <- puzzles %>% correlation_qq(x = SolveTime, y = MindPractice) + labs(x="", title = "Mind Practice")
p2 <- puzzles %>% correlation_qq(x = SolveTime, y = SleepHours) + labs(x="", y="", title = "Sleep Hours")
p3 <- puzzles %>% correlation_qq(x = SolveTime, y = FocusMinutes) + labs(title = "Focus Minutes")
p4 <- puzzles %>% correlation_qq(x = SolveTime, y = PuzzleComplexity) + labs(y="",title = "Puzzle Complexity")

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```

## Example 

- Because we met the normality assumption, we can use Pearson's correlation to quantify pairwise relationships.

```{r}
#| echo: false
puzzles %>% correlation() %>% filter(var1 == "SolveTime" | var2 == "SolveTime") 
```

## Example 

- Because we met the normality assumption, we can use Pearson's correlation to quantify pairwise relationships.

```{r}
#| echo: false
puzzles %>% correlation(x = SolveTime, y = MindPractice) 
```

- Mind practice is significantly correlated with solve time ($p < 0.001$).
    - The correlation is negative and modest ($r \approx -0.42$), indicating that increased mind practice is associated with decreased solve times (i.e., they are faster to solve).

## Example 

- Because we met the normality assumption, we can use Pearson's correlation to quantify pairwise relationships.

```{r}
#| echo: false
puzzles %>% correlation(x = SolveTime, y = SleepHours) 
```

- Sleep duration is significantly correlated with solve time ($p < 0.001$).
    - The correlation is negative and weak ($r \approx -0.27$), indicating that increased sleep duration is associated with decreased solve times (i.e., they are faster to solve).
    
## Example 

- Because we met the normality assumption, we can use Pearson's correlation to quantify pairwise relationships.

```{r}
#| echo: false
puzzles %>% correlation(x = SolveTime, y = FocusMinutes) 
```

- Focus is not significantly correlated with solve time ($p = 0.219$).
    - The correlation is negative and weak ($r \approx -0.10$), indicating that increased focus is associated with decreased solve times (i.e., they are faster to solve).  
    
## Example 

- Because we met the normality assumption, we can use Pearson's correlation to quantify pairwise relationships.

```{r}
#| echo: false
puzzles %>% correlation(x = SolveTime, y = PuzzleComplexity) 
```

- Puzzle complexity is significantly correlated with solve time ($p < 0.001$).
    - The correlation is positive and modest ($r \approx 0.64$), indicating that increased complexity is associated with increased solve times (i.e., they are slower to solve).    

## Example 

- Let's now examine the multiple regression model,

```{r}
#| echo: false
puzzles %>% linear_regression(outcome = SolveTime,
                           function_of = MindPractice + SleepHours + FocusMinutes + PuzzleComplexity)
```

- This gives the regression model,

$$
\hat{\text{solve time}} = 37.54 - 4.12 \text{ practice} - 1.82 \text{ sleep} - 0.02 \text{ focus} + 3.10 \text{ complexity}
$$

## Example 

- If I were reporting the table of results, I would submit

| Predictor         | Estimate (95% CI)            | p-value |
|:------------------|:----------------------------:|--------:|
| Mind Practice     | -4.12 (-5.02, -3.22)        | <0.001  |
| Sleep Hours      | -1.82 (-2.58, -1.06)        | <0.001  |
| Focus Minutes     | -0.02 (-0.21, 0.17)       | 0.857   |
| Puzzle Complexity | 3.10 (2.63, 3.57)          | <0.001  |

## Example 

- Before we make inference on each predictor, let's examine the significance of the regression model.

```{r}
#| echo: false
puzzles %>% significant_line(outcome = SolveTime,
                           function_of = MindPractice + SleepHours + FocusMinutes + PuzzleComplexity)
```

## Example 

- **Hypotheses**
    - $H_0: \beta_{\text{mind}} = \beta_{\text{sleep}} = \beta_{\text{focus}} = \beta_{\text{complex}} = 0$ 
    - $H_1:$ at least one $\beta_i \ne 0$

- **Test Statistic and *p*-Value**
    - $\chi^2_0 = 8661.51$; $p < 0.001$

- **Rejection Region**
    - Reject $H_0$ if $p<\alpha$; $\alpha=0.05$
    
- **Conclusion and Intepretation**
    - Reject $H_0$. There is sufficient evidence to conclude that at least one of the predictors is significantly associated with solve time, after adjusting for the other predictors.

## Example 

- Now, let's provide some interpretations and comment on significance.

| Predictor         | Estimate (95% CI)            | p-value |
|:------------------|:----------------------------:|--------:|
| Mind Practice     | -4.12 (-5.02, -3.22)        | <0.001  |
| Sleep Hours      | -1.82 (-2.58, -1.06)        | <0.001  |
| Focus Minutes     | -0.02 (-0.21, 0.17)       | 0.857   |
| Puzzle Complexity | 3.10 (2.63, 3.57)          | <0.001  |

- Mind practice is significantly associated with solve time ($p < 0.001$). After adjusting for sleep, focus, and complexity, each additional hour of mind practice is associated with a decrease in solve time of approximately 4.12 minutes (95% CI: -5.02, -3.22).

## Example 

- Now, let's provide some interpretations and comment on significance.

| Predictor         | Estimate (95% CI)            | p-value |
|:------------------|:----------------------------:|--------:|
| Mind Practice     | -4.12 (-5.02, -3.22)        | <0.001  |
| Sleep Hours      | -1.82 (-2.58, -1.06)        | <0.001  |
| Focus Minutes     | -0.02 (-0.21, 0.17)       | 0.857   |
| Puzzle Complexity | 3.10 (2.63, 3.57)          | <0.001  |

- Sleep hours is significantly associated with solve time ($p < 0.001$). After adjusting for mind practice, focus, and complexity, each additional hour of sleep is associated with a decrease in solve time of approximately 1.82 minutes (95% CI: -2.58, -1.06).

## Example 

- Now, let's provide some interpretations and comment on significance.

| Predictor         | Estimate (95% CI)            | p-value |
|:------------------|:----------------------------:|--------:|
| Mind Practice     | -4.12 (-5.02, -3.22)        | <0.001  |
| Sleep Hours      | -1.82 (-2.58, -1.06)        | <0.001  |
| Focus Minutes     | -0.02 (-0.21, 0.17)       | 0.857   |
| Puzzle Complexity | 3.10 (2.63, 3.57)          | <0.001  |

- Focus minutes is not significantly associated with solve time ($p = 0.857$). After adjusting for mind practice, sleep, and complexity, each additional minute of focus is associated with a decrease in solve time of approximately 0.02 minutes (95% CI: -0.21, 0.17).

## Example 

- Now, let's provide some interpretations and comment on significance.

| Predictor         | Estimate (95% CI)            | p-value |
|:------------------|:----------------------------:|--------:|
| Mind Practice     | -4.12 (-5.02, -3.22)        | <0.001  |
| Sleep Hours      | -1.82 (-2.58, -1.06)        | <0.001  |
| Focus Minutes     | -0.02 (-0.21, 0.17)       | 0.857   |
| Puzzle Complexity | 3.10 (2.63, 3.57)          | <0.001  |

- Puzzle complexity is significantly associated with solve time ($p < 0.001$). After adjusting for mind practice, sleep, and focus, each additional unit increase in puzzle complexity is associated with an increase in solve time of approximately 3.10 minutes (95% CI: 2.63, 3.57).

## Example 

- Let's create a visualization of the model.
    
    - We know that solve time will be on the *y*-axis.
    - We can choose one predictor to put on the *x*-axis -- let's choose sleep duration.
    - The other predictors will be set to their median values.
    
```{r}
#| echo: false
puzzles %>% mean_median(everything())
```

## Example 

- Our first step is to create predicted values using the found regression model.
    - Let sleep duration vary.
    - Plug in medians for other predictors.
    
$$
\hat{\text{solve time}} = 37.54 - 4.12 \text{ practice} - 1.82 \text{ sleep} - 0.02 \text{ focus} + 3.10 \text{ complexity}
$$    

```{r}
puzzles <- puzzles %>%
  mutate(predicted = 37.54 - 4.12 * median(MindPractice)
                           - 1.82 * SleepHours
                           - 0.02 * median(FocusMinutes)
                           + 3.10 * median(PuzzleComplexity))
```

## Example 

- Then we can graph.
    - Scatterplot: *x* = SleepHours, *y* = SolveTime
    - Line: *x* = SleepHours, *y* = predicted

```{r}
#| eval: false
puzzles %>% ggplot(aes(x = SleepHours, y = SolveTime)) + # specify x and y
  geom_point(size = 2.5, color = "gray50") + # plot points
  geom_line(aes(y = predicted), linewidth = 1, color = "black") + # plot line
  labs(x = "Sleep Duration (hours)", # edit x-axis label
       y = "Time to Complete Puzzle (min)") + # edit y-axis label
  theme_bw() # change theme of graph
```

## Example 

- The resulting graph,

```{r}
#| echo: false
#| fig-align: 'center'
puzzles %>% ggplot(aes(x = SleepHours, y = SolveTime)) + # specify x and y
  geom_point(size = 2.5, color = "gray50") + # plot points
  geom_line(aes(y = predicted), linewidth = 1, color = "black") + # plot line
  labs(x = "Sleep Duration (hours)", # edit x-axis label
       y = "Time to Complete Puzzle (min)") + # edit y-axis label
  theme_bw() # change theme of graph
```

## Wrap Up

- This module covers the basics of linear regression.

- As a reminder, this is just the beginning. There are many more advanced topics to explore, including:

    - Interaction terms
    - Nonlinear relationships
    - Model selection
    - Diagnostics and remedial measures
    
- STA4231 - Statistics for Data Science II dives deeper into regression topics.
