---
title: "Module 4 Review"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    theme: default
    css: mlp2.css
    self-contained: true
    slide-number: true
    footer: "STA4173 - Biostatistics - Fall 2025"
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Putting It All Together

```{r}
#| echo: false
library(tidyverse)
library(ssstats)
```

- Relationships between categorical variables.

- One-sample proportion.
- Two-sample proportions.
- Test for goodness-of-fit.
    - Uniform distribution.
    - Historical distribution.
- Test for Independence.    

## Example 1: One-Sample Proportions

- Rudolph surveys elves in the North Pole Workshop to see if they think his nose is "bright enough for foggy nights." He believes at least 90% of elves think so, and he wants to test that claim.
    
```{r}
#| echo: false
library(tidyverse)
library(ssstats)

set.seed(1225)

# true underlying proportion (you can tweak this)
true_p <- 0.85   # a bit less than 0.90, so we may reject H0
n_elves <- 120

rudolph <- data.frame(
  elf_id = 1:n_elves,
  nose_bright = rbinom(n_elves, size = 1, prob = true_p)
)

# make it pretty for students (Yes/No)
rudolph <- rudolph |>
  dplyr::mutate(
    nose_bright = ifelse(nose_bright == 1, "Yes", "No")
  )
```

```{r}
rudolph %>% n_pct(nose_bright)
```
        
## Example 1: One-Sample Proportions

- Rudolph surveys elves in the North Pole Workshop to see if they think his nose is "bright enough for foggy nights." He believes at least 80% of elves think so, and he wants to test that claim.

- Let's first look at the 95% confidence interval for $\pi$, the proportion of elves who think Rudolph's nose is bright enough.

```{r}
rudolph %>% one_prop_CI(binary = nose_bright,
                        event = "Yes")
```

- Thus, the 95% CI for $\pi$, the true proportion of elves who think Rudolph's nose is bright enough for foggy nights, is (0.80, 0.92).

## Example 1: One-Sample Proportions

- Rudolph surveys elves in the North Pole Workshop to see if they think his nose is "bright enough for foggy nights." He believes at least 90% of elves think so, and he wants to test that claim.

- Now, let's test the hypothesis -- do at least 85% of elves think Rudolph's nose is bright enough? Test at the $\alpha=0.05$ level.

```{r}
rudolph %>% one_prop_HT(binary = nose_bright,
                        event = "Yes",
                        p = 0.8,
                        alternative = "greater")
```

## Example 1: One Sample Proportions

- **Hypotheses:**
    - $H_0: \ \pi \le 0.80$
    - $H_1: \ \pi > 0.80$

- **Test Statistic and *p*-Value**
    - $z_0 = 2.05$, $p = 0.020$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$ ($0.02 < 0.05$). There is sufficient evidence to suggest that at least 80% of elves believe Rudolph's nose is bright enough for foggy nights.
    
## Example 2: Two Sample Proportions

- Rudolph proposes backup LED antlers for safety. He surveys female and male reindeer about whether they support the LED plan. Rudolph wants to test whether support is higher among female reindeer than male reindeer.

```{r}
#| echo: false
set.seed(1226)

n_female <- 40
n_male   <- 50

# true underlying support probabilities
p_female <- 0.75
p_male   <- 0.55

rudolph <- dplyr::tibble(
  reindeer_id = 1:(n_female + n_male),
  sex = c(rep("Female", n_female), rep("Male", n_male)),
  support_led = c(
    rbinom(n_female, size = 1, prob = p_female),
    rbinom(n_male,   size = 1, prob = p_male)
  )
) |>
  dplyr::mutate(
    support_led = ifelse(support_led == 1, "Support", "Do not support")
  )

```

```{r}
rudolph %>% n_pct(support_led, sex)
```

## Example 2: Two Sample Proportions

- Rudolph proposes backup LED antlers for safety. He surveys female and male reindeer about whether they support the LED plan. Rudolph wants to test whether support is higher among female reindeer than male reindeer.

- Let's find the 90% CI for $\pi_{\text{F}}-\pi_{\text{M}}$.

```{r}
rudolph %>% two_prop_CI(binary = support_led,
                        grouping = sex,
                        event = "Support",
                        confidence = 0.90)
```

## Example 2: Two Sample Proportions

- Rudolph proposes backup LED antlers for safety. He surveys female and male reindeer about whether they support the LED plan. Rudolph wants to test whether support is higher among female reindeer than male reindeer.

- Let's now determine if support is indeed higher among female reindeer. Test at the $\alpha=0.10$ level.

```{r}
#| eval: false
rudolph %>% two_prop_HT(binary = support_led,
                        grouping = sex,
                        event = "Support",
                        alternative = "greater",
                        alpha = 0.1)
```

## Example 2: Two Sample Proportions

```{r}
rudolph %>% two_prop_HT(binary = support_led,
                        grouping = sex,
                        event = "Support",
                        alternative = "greater",
                        alpha = 0.1)
```

## Example 2: Two Sample Proportions

- **Hypotheses:**
    - $H_0: \ \pi_{\text{F}} \le \pi_{\text{M}}$
    - $H_1: \ \pi_{\text{F}} > \pi_{\text{M}}$

- **Test Statistic and *p*-Value**
    - $z_0 = 1.43$, $p = 0.076$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.10$
    
- **Conclusion and interpretation**
    - Reject $H_0$ ($0.07 < 0.10$). There is sufficient evidence to suggest that support for LED antlers is higher among female reindeer than male reindeer.
    
## Example 3: Goodness of Fit (Uniform)   

- In the North Pole Wrapping Department, Rudolph wonders whether children are equally likely to request each of six magical gift wrap colors: red, green, blue, yellow, pink, or gold.

```{r}
#| echo: false

set.seed(1230)

# Number of gift requests
n_kids <- 300

wrap_colors <- c("Red", "Green", "Blue", "Yellow", "Pink", "Gold")

# TRUE underlying probabilities (slightly non-uniform)
# You can switch this to rep(1/6, 6) if you want perfectly uniform data
true_probs <- c(0.15, 0.18, 0.17, 0.16, 0.18, 0.16)

# Simulate the dataset
rudolph <- tibble(
  kid_id = 1:n_kids,
  wrap_color = sample(wrap_colors, size = n_kids, replace = TRUE, prob = true_probs)
)
```

```{r}
rudolph %>% n_pct(wrap_color, rows = 6)
```

## Example 3: Goodness of Fit (Uniform)   

- In the North Pole Wrapping Department, Rudolph wonders whether children are equally likely to request each of six magical gift wrap colors: red, green, blue, yellow, pink, or gold. Let's test at the $\alpha=0.05$ level.

```{r}
rudolph %>% goodness_of_fit(categorical = wrap_color,
                           alpha = 0.05)
```

## Example 3: Goodness of Fit (Uniform)   

- **Hypotheses:**
    - $H_0:$ $\pi_{\text{R}} = pi_{\text{Gr}} = \pi_{\text{B}} = \pi_{\text{Y}} = \pi_{\text{P}} = \pi_{\text{Go}} = \frac{1}{6}$
    - $H_1:$ At least one $\pi_i \ne \frac{1}{6}$.

- **Test Statistic and *p*-Value**
    - $\chi^2_0 = 2.08$, $p = 0.838$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$ ($0.838 > 0.05$). There is not sufficient evidence to suggest that the proportions of gift wrap color are unequal.

## Example 4: Goodness of Fit (Historical Distribution)

- Santa's forecasting model predicts the distribution of toy requests:

    - Trains: 30%
    - Dolls: 25%
    - Teddy Bears: 20%
    - Games: 25%

- Rudolph samples 400 letters to Santa and records the most requested toy per letter in order to determine if the observed distribution matches Santa's forecast.

```{r}
#| echo: false

set.seed(18671)

n_letters <- 400
toy_types <- c("Trains", "Dolls", "Teddy Bears", "Games")

# true probabilities (can match or deviate from Santa's model)
true_probs <- c(0.28, 0.27, 0.18, 0.27)   # slightly off from model

rudolph <- dplyr::tibble(
  letter_id = 1:n_letters,
  toy_request = sample(toy_types, size = n_letters, replace = TRUE, prob = true_probs)
)
```

```{r}
rudolph %>% n_pct(toy_request, rows = 4)
```

## Example 4: Goodness of Fit (Historical Distribution)

- Santa's forecasting model predicts the distribution of toy requests:

    - Trains: 30%
    - Dolls: 25%
    - Teddy Bears: 20%
    - Games: 25%

- Rudolph samples 400 letters to Santa and records the most requested toy per letter in order to determine if the observed distribution matches Santa's forecast. Because we want to be super sure, we will test at the $\alpha=0.01$ level.

```{r}
#| eval: false
rudolph %>% goodness_of_fit(categorical = toy_request,
                           expected = c("Trains" = 0.30, 
                                        "Dolls" = 0.25, 
                                        "Teddy Bears" = 0.20, 
                                        "Games" = 0.25),
                           alpha = 0.01)
```

## Example 4: Goodness of Fit (Historical Distribution)

```{r}
rudolph %>% goodness_of_fit(categorical = toy_request,
                           expected = c("Trains" = 0.30, 
                                        "Dolls" = 0.25, 
                                        "Teddy Bears" = 0.20, 
                                        "Games" = 0.25),
                           alpha = 0.01)
```

## Example 4: Goodness of Fit (Historical Distribution)

- **Hypotheses:**
    - $H_0:$ The proportions do not differ from Santa's expected distribution.
    - $H_1:$ The proportions differ from Santa's expected distribution.

- **Test Statistic and *p*-Value**
    - $\chi^2_0 = 18.19$, $p < 0.001$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.01$
    
- **Conclusion and interpretation**
    - Reject $H_0$ ($(<0.001) < 0.10$). There is sufficient evidence to suggest that the distribution of toy requests differs from Santa's expected distribution.

## Example 5: Test for Independence

```{r}
#| echo: false
set.seed(964568)

n_reindeer <- 120

team_levels <- c("Lead", "Middle", "Back")

# Probability of being in each team position (just for simulation)
team_probs <- c(0.2, 0.5, 0.3)

reindeer_team <- sample(team_levels, size = n_reindeer, replace = TRUE, prob = team_probs)

misfit_levels <- c("Jack-in-the-Box", "Elephant with Polka Dots", "Charlie-in-the-Box")

# Conditional probabilities for toy preference given team
# Each row corresponds to a team, each column to a Misfit Toy
toy_prob_matrix <- rbind(
  Lead   = c(0.50, 0.30, 0.20),  # Leads like Jack more
  Middle = c(0.30, 0.40, 0.30),
  Back   = c(0.20, 0.30, 0.50)   # Backs like Charlie more
)

get_toy_pref <- function(team) {
  probs <- toy_prob_matrix[team, ]
  sample(misfit_levels, size = 1, prob = probs)
}

misfit_pref <- sapply(reindeer_team, get_toy_pref)

rudolph <- dplyr::tibble(
  reindeer_id = 1:n_reindeer,
  team = factor(reindeer_team, levels = team_levels),
  misfit_toy = factor(misfit_pref, levels = misfit_levels)
)
```

- Rudolph suspects that a reindeer's team position (lead, middle, back) might be associated with which Misfit Toy they identify with most: Jack-in-the-Box, Elephant with Polka Dots, or Charlie-in-the-Box. 

- After surveying 120 reindeer, Rudolph wants to test if favorite Misfit Toy depends on team position.

```{r}
rudolph %>% n_pct(misfit_toy, team)
```

## Example 5: Test for Independence

- Rudolph suspects that a reindeer's team position (lead, middle, back) might be associated with which Misfit Toy they identify with most: Jack-in-the-Box, Elephant with Polka Dots, or Charlie-in-the-Box. 

- After surveying 120 reindeer, Rudolph wants to test if favorite Misfit Toy depends on team position. Test at the $\alpha=0.05$ level.

```{r}
rudolph %>% independence_test(var1 = misfit_toy,
                              var2 = team,
                              alpha = 0.05)
```

## Example 5: Test for Independence 

- **Hypotheses:**
    - $H_0:$ A reindeer's favorite misfit toy does not depend on team position.
    - $H_1:$ A reindeer's favorite misfit toy depends on team position.

- **Test Statistic and *p*-Value**
    - $\chi^2_0 = 15.85$, $p = 0.003$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$ ($0.003 < 0.05$). There is sufficient evidence to suggest that a reindeer's favorite misfit toy depends on team position.

## Wrap Up

- This module covers the basics of categorical analysis.

- Key topics include:
    - One-sample proportion inference.
    - Two-sample proportion inference.
    - Chi-square tests for goodness-of-fit (uniform and historical).
    - Chi-square test for independence.
    
- Note that we could extend what we know about regression (using continuous outcomes) to categorical outcomes.
    - i.e., we are not limited by the methods in this module.
